---
title: "Covid-19 and regional deprivation in England"
author: "Dr Matthew Gwynfryn Thomas"
date: "18/02/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading, include=FALSE}
library(tidyverse)

# Load files created by `prep data.R`
deaths_msoa <- read_csv("data/deaths-msoa.csv")
deaths_la <- read_csv("data/deaths-la.csv")

# Regression slopes from `analyse deprivation.R`
slopes_extent <- read_csv("data/regression-slopes.csv")
slopes_score <- read_csv("data/regression-slopes-scores.csv")
```

## Deaths and deprivation in Local Authorities
Overall, there's a strong association between Covid-19 death rates and deprivation - more deprived Local Authorities have higher death rates:

```{r la_deaths_overall, echo=FALSE, message=FALSE, warning=FALSE}
deaths_la %>% 
  ggplot(aes(y = DeathRate, x = Score, colour = RGN19NM, fill = RGN19NM)) +
  geom_point() +
  geom_smooth(method = "lm", colour = "black", fill = "grey") +
  labs(y = "Death rate per 100,000 population", x = "IMD - average score (higher is more deprived)", colour = NULL, fill = NULL) +
  theme_classic() +
  theme(legend.position = "bottom")
```

Generally, Covid-19 death rates are higher in more deprived Local Authorities, regardless of region - except in London and the South West - although the strength of the association varies from place to place:

```{r la_deaths_score, echo=FALSE, message=FALSE, warning=FALSE}
deaths_la %>% 
  ggplot(aes(y = DeathRate, x = Score, colour = RGN19NM, fill = RGN19NM)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~RGN19NM) +
  labs(y = "Death rate per 100,000 population", x = "IMD - average score (higher is more deprived)", colour = NULL, fill = NULL) +
  theme_classic() +
  theme(legend.position = "bottom")
```

And the pattern is broadly the same when looking at proportions of the population who live within the nation's most deprived neighbourhoods, rather than just the Local Authorities' average deprivation scores:

```{r la_deaths_extent, echo=FALSE, message=FALSE, warning=FALSE}
deaths_la %>% 
  ggplot(aes(y = DeathRate, x = Extent, colour = RGN19NM, fill = RGN19NM)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~RGN19NM) +
  scale_x_continuous(labels = scales::percent) +
  labs(y = "Death rate per 100,000 population", x = "Percent of population living in England's most deprived areas", colour = NULL, fill = NULL) +
  theme_classic() +
  theme(legend.position = "bottom")
```

## MSOA-level deprivation
Similarly, death rates tend to be higher in more deprived neighbourhoods, although the association isn't as strong compared to looking at Local Authorities:

```{r msoa_deaths_overall, echo=FALSE, message=FALSE, warning=FALSE}
deaths_msoa %>% 
  ggplot(aes(y = DeathRate, x = Score, colour = RGN19NM, fill = RGN19NM)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", colour = "black", fill = "grey") +
  labs(y = "Death rate per 100,000 population", x = "IMD - average score (higher is more deprived)", colour = NULL, fill = NULL) +
  theme_classic() +
  theme(legend.position = "bottom")
```

However, when looking at neighbourhood-level deprivation by region, the association between death rates and deprivation seems a lot weaker:

```{r msoa_deaths_score, echo=FALSE, message=FALSE, warning=FALSE}
deaths_msoa %>% 
  ggplot(aes(y = DeathRate, x = Score, colour = RGN19NM, fill = RGN19NM)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", colour = "black", fill = "grey") +
  facet_wrap(~RGN19NM) +
  labs(y = "Death rate per 100,000 population", x = "IMD - average score (higher is more deprived)", colour = NULL, fill = NULL) +
  theme_classic() +
  theme(legend.position = "bottom")
```

And this pattern remains similar when using other summary measures of deprivation, such as the proportion of the population who live within the nation's most deprived neighbourhoods...

```{r msoa_deaths_extent, echo=FALSE, message=FALSE, warning=FALSE}
deaths_msoa %>% 
  ggplot(aes(y = DeathRate, x = Extent, colour = RGN19NM, fill = RGN19NM)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", colour = "black", fill = "grey") +
  facet_wrap(~RGN19NM) +
  scale_x_continuous(labels = scales::percent) +
  labs(y = "Death rate per 100,000 population", x = "Percent of population living in England's most deprived areas", colour = NULL, fill = NULL) +
  theme_classic() +
  theme(legend.position = "bottom")
```

... or the proportion of LSOAs within an MSOA which are among the nation's most deprived:

```{r msoa_deaths_proportion, echo=FALSE, message=FALSE, warning=FALSE}
deaths_msoa %>% 
  ggplot(aes(y = DeathRate, x = Proportion, colour = RGN19NM, fill = RGN19NM)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", colour = "black", fill = "grey") +
  facet_wrap(~RGN19NM) +
  scale_x_continuous(labels = scales::percent) +
  labs(y = "Death rate per 100,000 population", x = "Percent of neighbourhoods in England's most deprived 10%", colour = NULL, fill = NULL) +
  theme_classic() +
  theme(legend.position = "bottom")
```

So although more deprived areas have higher death rates on average at a national level and Local Authority level, the picture becomes a bit blurred when looking at neighbourhoods.

## Modelling regional deprivation and death rates
For Local Authorities, higher deprivation is associated with higher death rates in the East of England, South East, West Midlands, and Yorkshire and the Humber.

At MSOA-level, higher deprivation is associated with higher death rates in the North East, North West, and Yorkshire and the Humber. 

```{r slopes_extent, echo=FALSE, message=FALSE, warning=FALSE}
slopes_extent %>% 
  ggplot(aes(x = Region, y = Q50, group = Model, colour = Model, fill = Model)) +
  geom_pointrange(aes(ymin = Q2.5, ymax = Q97.5), position = position_dodge(width = 0.4)) +
  geom_hline(yintercept = 0, lty = 2, colour = "grey") +
  labs(x = NULL, y = "Predicted death rate in highest-deprivation areas",
       fill = NULL, colour = NULL,
       title = "Predicted change in death rates for highest-deprivation areas compared to lowest",
       subtitle = "Dots are posterior medians; lines are 95% credible intervals") +
  theme_classic() +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1)  # rotate x-axis labels
  )
```

These results are from a Bayesian multilevel regressions modelling how extent of deprivation predicts Covid-19 death rates in LAs and MSOAs, with varying intercepts and varying slopes for regions.

## Which measure of deprivation you use matters
The results above are based on the proportion of the population who live within the nation's most deprived neighbourhoods. If we instead model death rates using average deprivation scores, a slightly different picture emerges. Now there is also a statistically significant association between deprivation and death rates in the East and West Midlands for MSOAs. At LA-level, significant associations are found in the East Midlands, Yorkshire and the Humber, as well as the East and South East.

```{r slopes_score, echo=FALSE, message=FALSE, warning=FALSE}
slopes_score %>% 
  ggplot(aes(x = Region, y = Q50, group = Model, colour = Model, fill = Model)) +
  geom_pointrange(aes(ymin = Q2.5, ymax = Q97.5), position = position_dodge(width = 0.4)) +
  geom_hline(yintercept = 0, lty = 2, colour = "grey") +
  labs(x = NULL, y = "Predicted death rate",
       fill = NULL, colour = NULL,
       title = "Predicted death rates given a one standard deviation change in deprivation scores",
       subtitle = "Dots are posterior medians; lines are 95% credible intervals") +
  theme_classic() +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1)  # rotate x-axis labels
  )
```
